$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-05 paper/paper.tex
Minor changes
more on leftovers

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
\todo{rename gradual to something else...}
\todo{replace judgment with judgement everywhere!}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context $\PreCtx_n$ and (input) usage context $\Ctx_{idxs}$, adds an extra {output usage context} to the typing rules.

This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgement with both an input and an output usage context.
  \item
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can be stated thanks to the presence of a leftover context.
\end{itemize}
  \todo{I still don't get the last point above, as weakening and strengthening were already possible to state in \picalc{} before. What do you mean with can be stated here?}

Our type system with leftovers is composed of two typing relations: one for variable references (\autoref{def:varref}) and one for processes (\autoref{def:types}).
Both relations are indexed by a typing context $\gamma$, an input usage context $\Gamma$, and an output usage context $\Delta$ (containing the leftovers).

The typing judgement for variables $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ asserts that ``index $i$ is of type $t$ in $\gamma$, and subtracting $y$ at position $i$ from $\Gamma$ results in $\Delta$''.
The typing judgement for processes $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$, which are the leftovers.
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{functionality} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{nidefinition}

@@

  The inaction process does not change usage annotations.
  Scope restriction expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ that is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that these are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

  Receiving needs a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.

  Similar to input, sending needs a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ of the resources it uses, we save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

\begin{example}[Typing derivation]
  \label{example-derivations}
  Let us type a derivation of the process $\func{p}$ using linear and shared types. 
  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transiting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process is typed following its syntax.
  On the right-hand, we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmited.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\todo{highlight that we can talk about a resource even if it has been exhausted (e.g. we can send 0 multiplicities of a channel over another channel}
We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.
\todo{rename gradual to something else...}
\todo{replace judgment with judgement everywhere!}
\todo{Theorem to Thm. Definiton to Def. when referencing.}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our type system, an approach that, in addition to the usual typing context $\PreCtx_n$ and (input) usage context $\Ctx_{idxs}$, adds an extra \emph{output usage context} to the typing rules.

This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
Leftover typing presents three main advantages:
\begin{itemize}
  \item \textbf{Encapsulation}:
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgement with both an input and an output usage context.
  \item \textbf{No context split}:
    No need for an extensional context splitting proof for parallel composition.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item \textbf{New theorems}:
    For the first time, theorems like \textbf{framing} (\autoref{thm:framing}), \textbf{weakening} (\autoref{thm:weakening}) and \textbf{strengthening} (\autoref{thm:strengthening}) can now be stated within the linear \picalc{} --- as well as gradual and shared, thanks to the presence of a leftover context.
\end{itemize}

Our type system with leftovers is composed of two typing relations: one for variable references (\autoref{def:varref}) and one for processes (\autoref{def:types}).
Both relations are indexed by a typing context $\gamma$, an input usage context $\Gamma$, and an output usage context $\Delta$ (the leftovers).

The typing judgement for variables $\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta}$ asserts that ``index $i$ is of type $t$ in typing context $\gamma$, and subtracting $y$ at position $i$ from input usage context $\Gamma$ results in leftover $\Delta$''.
The typing judgement for processes $\types{\gamma}{\Gamma}{P}{\Delta}$ asserts that ``process $P$ is well-typed under typing context $\gamma$, usage input context $\Gamma$ and leftovers $\Delta$''.

@@

  The base case $\constr{0}$ splits the usage annotation $x$ of type $\Usage_{idx}$ into $y$ and $z$ (the leftovers).
  This splitting $\opsquared{x}{y}{z}$ is as per the usage algebra provided by the developer for the index $idx$.
  We use the \emph{functionality} of the monoidal relation to alleviate the user from the proof burden $\opsquared{x}{y}{z}$.
  The inductive case $\suc$ appends the type $t'$ to the typing context, and the usage annotation $x'$ to both the input and output usage contexts. 
\end{nidefinition}

@@

  The inaction process in rule $\constr{end}$ does not change usage annotations.
  The scope restriction in rule $\constr{chan}$ expects three arguments: the type $t$ of data being transmitted; the usage annotation $x$ that is being transmitted; and the multiplicity $y$ given to the channel itself.
  This multiplicity $y$ is used for both input and output, so that these are balanced.
  The continuation process $P$ is provided with the new channel with usage annotation $(y \comma y)$, which it must completely exhaust.

  The input process in rule $\constr{recv}$ requires a channel $chan_i$ at index $i$ with usage $\li$ available, such that data with type $t$ and usage $x$ can be sent over it.
  Note that the index $i$ is used in the syntax of the typed process.
  We use the leftovers $\Xi$ to type the continuation process, which is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{0}$.
  The received element $x$ must be completely exhausted by the continuation process.

  Similarly to input, the output processes in rule $\constr{send}$ requires a channel $chan_i$ at index $i$ with usage $\lo$ available, such that data with type $t$ and usage $x$ can be sent over it.
  We use the leftover context $\Delta$ to type the continuation process, which needs an element $loc_j$ at index $j$ with type $t$ and usage $x$, as per the type of channel $chan_i$.
  The leftovers $\Xi$ are used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition in rule $\constr{comp}$ uses the leftovers of the left-hand process to type the right-hand process.
  By keeping track in the typing derivation of $P$ the resources it uses, we save the user from having to provide an extrinsic proof of context split.
\end{nidefinition}

\begin{example}[Typing derivation]
  \label{example-derivations}
  We will now illustrate a typing derivation of the process $\func{p}$ using linear and shared types, given in line ?? of Listing \autoref{}.

  Let $\func{\omega}$ be a shortcut for $(\one \comma \one)$ in an algebra of shared types.
  Let $\epsilon$ be the empty usage context --- where all usage annotations are $(\zero \comma \zero)$.
  To type $\func{p}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $(\one \comma \one)$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we will only transmit its input capability.
  The left-hand process (line ??) is typed following its syntax.
  On the right-hand (line ??), we need the channel that is going to be sent.
  The type $\unit$ of this channel is entirely determined by the type of the channel over which it is going to be transmitted.
  Its usage annotation is $(\one \comma \one)$: we will transmit $\li$ and save $\lo$ for the continuation process.
